Narrative

### 2023-05-20 manuscript/introduction.qmd
intro: minor edits final paragraph

Meta-analytic estimates are highly dependent on the quality of the data in the primary studies that the method summarizes. However, there is concern that researchers in psychology tend to neglect proper measurement (ref measurement-schmesurement) to the extent that psychology can be said to be in a measurement crisis (Kane et al., 2021). For example, Flake et al. (2017) report that although latent constructs are widely examined in personality and social psychology, researchers barely report any of the validity evidence required to ascertain the extent to which implemented scales measure the constructs of interest. Such inattention to measurement can also bias heterogeneity estimates. We will focus here on reliability of measurement (defined in classical test theory as the ratio of true score variance to measurement error variance; e.g., REF), which is the only measurement validity evidence researchers report consistently (Flake et al., 2017), and which can be corrected for in meta-analysis (Hunter & Schmidt; 2015).

@@

A negative bias can be expected in heterogeneity estimates of uncorrected effect sizes because in absolute terms larger effect sizes are more attenuated by a given reliability. That is, although the attenuation formula $r_{xy} = \rho_{xy} \times \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ implies that any effect size $\rho_{xy}$ will be attenuated by the same proportion given the same reliabilities $\sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$, this proportion corresponds to a larger absolute value for larger effect sizes. Consequently, in the presence of heterogeneity attenuation will move larger true effect sizes further towards zero than smaller ones, decreasing heterogeneity in observed effect sizes. This negative bias can be observed directly by assuming that an average effect size $\theta$ has some heterogeneity $\tau^2$, all studies have infinite sample size (i.e., no within-study sampling variance), and equal reliability ($R = R_i = R_{xx'} = R_{yy'}$). Any observed effect size $\hat{\theta_i}$ can then be seen as a study’s true effect size $\theta_i$ multiplied by the study reliability $R$. Hence, the distribution of observed effect sizes is the distribution of true effect sizes $N(\theta, \tau^2$ multiplied by the reliability $R$, such that $R \times N(\theta, \tau^2) = N(R \theta, R^2 \tau^2) = N(R \theta, R \tau)$. Because reliability lies between 0 – 1, under these conditions imperfect reliability implies that both the observed average effect size and the estimated heterogeneity will be less than their true value by a factor $\bar{R}$ (if heterogeneity is expressed as standard deviation rather than variance). This depiction is not entirely correct for Pearson's $r$ as it is bounded at {-1, 1}, but holds approximately if the mean effect size $\theta$ and heterogeneity $\tau^2$ are not too large. Hence, imperfect reliability in studies can be expected to create a negative bias in heterogeneity estimates whenever true heterogeneity is present. To make this concrete, consider an average reliability of 0.8 (the median reported reliability in psychology; Flake et al., 2017). Then, $0.8 \times N(\theta, \tau) = N(0.8\theta, 0.8 \tau)$, resulting in 20% less standard deviation among observed effect sizes than in true effect sizes. Table 2 illustrates this relationship for infinite sample size (i.e., no sampling variance) and three levels of reliability across three studies.

@@

It is difficult to predict the total effect of the negative and positive bias in heterogeneity due to unreliability, and existing literature tends to focus on the positive bias (e.g., Borenstein, 2009, p. 342; Card, 2015, p. 126; Hunter & Schmidt, 2015; Wiernick & Dahlke, 2020). However, given that the negative bias could often be 20% or more and the value of accurate  heterogeneity estimates for many research outcomes in psychology, we consider it important to have insight into the total bias in heterogeneity estimates that should be expected due to unreliability. This depends on average effect size, true heterogeneity, mean reliability across studies, the variability in reliabilities, and sampling variance within studies. Hence, we perform a Monte-Carlo simulation study to explore the expected bias in heterogeneity estimates when meta-analyzing primary studies with imperfect measurement reliabilities.

>>>>

Meta-analytic estimates are highly dependent on the quality of data in primary studies. However, there is concern that researchers in psychology tend to neglect proper measurement (Flake & Fried, 2020) to the extent that psychology can be said to be in a measurement crisis (Kane et al., 2021). For example, Flake et al. (2017) report that although latent constructs are widely examined in personality and social psychology, researchers barely report any of the validity evidence required to ascertain the extent to which implemented scales measure the constructs of interest. Such inattention to measurement can also bias heterogeneity estimates. We will focus here on reliability of measurement (defined in classical test theory as the ratio of true score variance to measurement error variance; e.g., REF), which is the only measurement validity evidence researchers report consistently (Flake et al., 2017), and which can be corrected for in meta-analysis (Hunter & Schmidt; 2015).

@@

A negative bias can be expected in heterogeneity estimates of uncorrected effect sizes because in absolute terms larger effect sizes are more attenuated by a given reliability. That is, although the attenuation formula $r_{xy} = \rho_{xy} \times \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ implies that any effect size $\rho_{xy}$ will be attenuated by the same proportion given the same reliabilities $\sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$, this proportion corresponds to a larger absolute value for larger effect sizes. Consequently, in the presence of heterogeneity attenuation will move larger true effect sizes further towards zero than smaller ones, decreasing heterogeneity in observed effect sizes. This negative bias can be observed directly by assuming an average effect size $\theta$ with some heterogeneity $\tau^2$, infinite sample size within studies (i.e., no within-study sampling variance), and equal reliability across studies ($R = R_i = R_{xx'} = R_{yy'}$). Any observed effect size $\hat{\theta_i}$ can then be seen as a study’s true effect size $\theta_i$ multiplied by the study reliability $R$. Hence, the distribution of observed effect sizes is the distribution of true effect sizes $N(\theta, \tau^2)$ multiplied by the reliability $R$, such that $R \times N(\theta, \tau^2) = N(R \theta, R^2 \tau^2) = N(R \theta, R \tau)$. Because reliability lies between 0 – 1, under these conditions imperfect reliability implies that both the observed average effect size and the estimated heterogeneity will be less than their true value by a factor $\bar{R}$ (if heterogeneity is expressed as standard deviation rather than variance). This depiction is not entirely correct for Pearson's $r$ as it is bounded at {-1, 1}, but holds approximately if the mean effect size $\theta$ and heterogeneity $\tau^2$ are not too large. Hence, imperfect reliability in studies can be expected to create a negative bias in heterogeneity estimates whenever true heterogeneity is present. Consider an average reliability of 0.8 (the median reported reliability in psychology; Flake et al., 2017). Then, $0.8 \times N(\theta, \tau) = N(0.8\theta, 0.8 \tau)$, resulting in 20% less standard deviation among observed effect sizes than in true effect sizes. Table 2 illustrates this relationship for infinite sample size (i.e., no sampling variance) and three levels of reliability across three studies.

@@

It is difficult to predict the total effect of the negative and positive bias in heterogeneity due to unreliability, and existing literature tends to focus on the positive bias (e.g., Borenstein, 2009, p. 342; Card, 2015, p. 126; Hunter & Schmidt, 2015; Wiernick & Dahlke, 2020). However, given the value of accurate  heterogeneity estimates for many research outcomes in psychology and that the negative bias could often be 20% or more, we consider it important for meta-analysts to have insight into the total bias in heterogeneity estimates that should be expected due to unreliability. This depends on average effect size, true heterogeneity, mean reliability across studies, the variability in reliabilities, and sampling variance within studies. Hence, we perform a Monte-Carlo simulation study to explore the expected bias in heterogeneity estimates when meta-analyzing primary studies with imperfect measurement reliabilities.
